### 2010-03-29 papers/icfp-2010-desc/paper_type_theory.tex
Desc@ICFP: get some sort of introduction.
Desc@ICFP: Tidied up the code in the paper, fixed a few transcription bugs and added colour (well, purple)

\section{The Type Theory}

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developed this technology in our
on-going work on Epigram~\cite{pigs:epigram}. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground-up.

@@

\[
\begin{array}{ll}
\Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
\Gamma \vdash \Bhab{t}{T}           & \mbox{term \(t\) has type \(T\) in context \(\Gamma\)} \\
\Gamma \vdash \Bhab{s \equiv t}{T}  & \mbox{\(s\) and \(t\) are equal at type \(T\) in context \(\Gamma\)} \\
\end{array}
\]

@@

\[
\begin{array}{l@{\;\Rightarrow\;\;}l}
\Gamma            \vdash \Bhab{t}{T}            
    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{T} \\
\Gamma            \vdash s \equiv \Bhab{t}{T}   
    & \Gamma \vdash \Bhab{s}{T} \;\wedge\; \Gamma\vdash \Bhab{t}{T} \\
\Gamma ; \xS ; \Delta \vdash J                      
    & \Gamma \vdash \Bhab{s}{S} \;\Rightarrow\; 
          \Gamma ; \Delta[s/x] \vdash J[s/x] \\
\end{array}
\]

@@

\begin{wstructure}
<- Context validity [figure]
    <- Not much to be said
\end{wstructure}

\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\Gamma       \vdash \Type{S}}
     {\Gamma ; \xS \vdash \Valid}\;x\not\in\Gamma
\]

\caption{Context validity}
\label{fig:context-validity}
\end{figure}

The rules ensuring Context Validity (Fig.~\ref{fig:context-validity})
are unsurprisingly common. They simply guarantee the validity of the
empty context, as well as the linear extension of a context by a
well-typed variable.

@@

The specification of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed of three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for diagnostic
purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
definition, we get that a finite set is a list of tags. Finally,
elements of a finite set $u$ belong to the corresponding $\EnumT{u}$
type. Intuitively, it corresponds to an index -- a number -- pointing
to an element of $u$.

@@

\[\stk{
%% spi
\spi{}{} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{e} \To \Set} \To \Set \\
\begin{array}{@{}ll@{\:\mapsto\:\:}l}
\spi{\NilE}{& P}        & \Unit \\
\spi{(\ConsE{t}{e})}{& P} & \TIMES{P\: \Ze}{\spi{e}{(\LAM{x} P\: (\Su{x}))}}
\end{array}
}\]

@@

\[\stk{
%% switch
\begin{array}{@{}ll}
\F{switch} : & \PITEL{\V{e}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{e} \To \Set} \\
             & \PITEL{\V{b}}{\spi{e}{P}}
               \PITEL{\V{x}}{\EnumT{e}} \To P~x
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\switch{(\ConsE{t}{e})}{P}{b}{\Ze}      & \fst{b} \\
\switch{(\ConsE{t}{e})}{P}{b}{(\Su{x})} & \switch{e}{(\LAM{x} P
  (\Su{x}))}{(\snd{b})}{x}
\end{array}
}\]

Again, there is a clear equivalent in the full-$\Set$ world: function
application. The operational behaviour of $\F{switch}$ is
straightforward: $x$ is peeled off as we move deeper inside the nested
tuple $b$. When $x$ equals $\Ze$, we simply return the value we are
pointing to.

@@

\[
\Rule{\Gamma ; \xS ; \Delta \vdash \Valid}
     {\Gamma ; \xS ; \Delta \vdash \Bhab{x}{S}}
\]

@@

\[
\Rule{\Gamma       \vdash \Type{S} \quad
      \Gamma ; \xS \vdash \Bhab{t}{T}}
     {\Gamma \vdash \Bhab{\PLAM{\x}{S} t}{\PIS{\xS} T}}
\]

@@

A first consequence is that we can actually denote elements of finite
sets \emph{by their name}. Indeed, such expression will be pushed into
a $\EnumT{u}$, where $\Bhab{u}{\EnumU}$ contains all the names
populating the finite set. It is therefore easy to elaborate a name
into the corresponding index.

@@

Finally, and more importantly, we are able to present our user an
abstraction of \emph{finite} function space. This work as follow. When
an expression $t$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.

>>>>

\section{The Type Theory}
\label{sec:type-theory}

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developed this technology in our
on-going work on Epigram. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground-up.

@@

\[
\begin{array}{ll}
\Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
\Gamma \vdash \Bhab{\V{t}}{\V{T}}           & \mbox{term \(\V{t}\) has type \(\V{T}\) in context \(\Gamma\)} \\
\Gamma \vdash \Bhab{\V{s} \equiv \V{t}}{\V{T}}  & \mbox{\(\V{s}\) and \(\V{t}\) are equal at type \(\V{T}\) in context \(\Gamma\)} \\
\end{array}
\]

@@

\[
\begin{array}{l@{\;\Rightarrow\;\;}l}
\Gamma            \vdash \Bhab{\V{t}}{\V{T}}            
    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{\V{T}} \\
\Gamma            \vdash \V{s} \equiv \Bhab{\V{t}}{\V{T}}   
    & \Gamma \vdash \Bhab{\V{s}}{\V{T}} \;\wedge\; \Gamma\vdash \Bhab{\V{t}}{\V{T}} \\
\Gamma ; \xS ; \Delta \vdash \V{J}                      
    & \Gamma \vdash \Bhab{\V{s}}{\V{S}} \;\Rightarrow\; 
          \Gamma ; \Delta[\V{s}/\V{x}] \vdash \V{J}[\V{s}/\V{x}] \\
\end{array}
\]

@@

\begin{wstructure}
<- Context validity [figure no longer]
    <- Not much to be said
\end{wstructure}

The rules ensuring Context Validity 
are unsurprisingly common. They simply guarantee the validity of the
empty context, as well as the linear extension of a context by a
well-typed variable.

\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\Gamma       \vdash \Type{\V{S}}}
     {\Gamma ; \xS \vdash \Valid}\;\V{x}\not\in\Gamma
\]

@@

The specification of finite sets is presented in
Figure~\ref{fig:typing-finite-set}. It is composed of three
components. First, we define tags as inhabitants of the $\UId$ type. A
tag is solely an informative token, used for diagnostic
purposes. Finite sets inhabits the $\EnumU$ type. Unfolding the
definition, we get that a finite set is a list of tags. Finally,
elements of a finite set $\V{u}$ belong to the corresponding $\EnumT{\V{u}}$
type. Intuitively, it corresponds to an index -- a number -- pointing
to an element of $\V{u}$.

@@

\[\stk{
%% spi
\spi{}{} : \PITEL{\V{e}}{\EnumU}
           \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \To \Set \\
\begin{array}{@{}ll@{\:\mapsto\:\:}l}
\spi{\NilE}{& \V{P}}        & \Unit \\
\spi{(\ConsE{\V{t}}{\V{e}})}{& \V{P}} & \TIMES{\V{P}\: \Ze}{\spi{\V{e}}{(\LAM{\V{x}} \V{P}\: (\Su{\V{x}}))}}
\end{array}
}\]

@@

\[\stk{
%% switch
\begin{array}{@{}ll}
\F{switch} : & \PITEL{\V{e}}{\EnumU}
               \PITEL{\V{P}}{\EnumT{\V{e}} \To \Set} \\
             & \PITEL{\V{b}}{\spi{\V{e}}{\V{P}}}
               \PITEL{\V{x}}{\EnumT{\V{e}}} \To P~x
\end{array} \\
\begin{array}{@{}l@{\:\mapsto\:\:}l}
\switch{(\ConsE{\V{t}}{\V{e}})}{\V{P}}{\V{b}}{\Ze}      & \fst{\V{b}} \\
\switch{(\ConsE{\V{t}}{\V{e}})}{\V{P}}{\V{b}}{(\Su{\V{x}})} & \switch{\V{e}}{(\LAM{\V{x}} \V{P}
  (\Su{\V{x}}))}{(\snd{\V{b}})}{\V{x}}
\end{array}
}\]

Again, there is a clear equivalent in the full-$\Set$ world: function
application. The operational behaviour of $\F{switch}$ is
straightforward: $\V{x}$ is peeled off as we move deeper inside the nested
tuple $\V{b}$. When $\V{x}$ equals $\Ze$, we simply return the value we are
pointing to.

@@

\[
\Rule{\Gamma ; \xS ; \Delta \vdash \Valid}
     {\Gamma ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
\]

@@

\[
\Rule{\Gamma       \vdash \Type{\V{S}} \quad
      \Gamma ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
     {\Gamma \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
\]

@@

A first consequence is that we can actually denote elements of finite
sets \emph{by their name}. Indeed, such expression will be pushed into
a $\EnumT{\V{u}}$, where $\Bhab{\V{u}}{\EnumU}$ contains all the names
populating the finite set. It is therefore easy to elaborate a name
into the corresponding index.

@@

Finally, and more importantly, we are able to present our user an
abstraction of \emph{finite} function space. This work as follow. When
an expression $\V{t}$ is pushed into a function which domain is a finite
set, we instead elaborate it as a term in a small $\pi$-type. That is,
a finite function. To realise our promise of a big $\Pi$-type, we
elaborate a call to the $\F{switch}$ eliminator.
